How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-22 Paper/paper.tex
minor changes

Our study is based on data from the Scratch project repository. By scraping the list of recent projects \footnote{\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed analysis on them. To the best of our knowledge, this is the first large-scale exploratory study on Scratch projects.

@@

We obtained the set of Scratch projects by using a scraping program. Our scraping program called Kragle starts by reading \url{https://scratch.mit.edu/explore/projects/all/} and thus obtaining project ids of projects that were most recently shared. Subsequently, Kragle obtains the JSON code for each of the listed projects.

@@

\subsection{Data analysis}
\label{dataAnalysis}
All scraped data, including the list of used blocks and parameters per project, was imported to a relational database. We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions. This data was then imported to spreadsheets for the calculation of statistics and the generation of graphical representations. We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects, lists of which we also provide as part of the dataset.\footnotemark[\ref{repo}]

@@

Apart from whole scripts we also examined cases where scripts differed only in the first (Hat) block.This way we examine if Scratch programmers assign the same functionality to handle different types of events. Cloned functionality blocks are found to be rare: without considering the first block, only 2,243 additional clones were found in 920 projects.

The second smell that we examine is the dead code smell. We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts. Investigating the first type, we find that a significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects. This is also shown in Figure \ref{fig:procedurecalls} and it occurs in 2,079 projects. For the second type, we examined the broadcast-receive messages and found that they are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received. This lack of syncronizations occurs in 18,669 projects (7.99\% of the total non-empty ones).

The third and fourth cases are detected in scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality. A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects. The majority of these scripts (86.6\%) are scripts missing the starting block. Examining the size of these dead scripts, 72.34\% are composed of a single block. As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}]

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/12deadcode}
	\caption{Size of scripts that are identified as dead code}
	\label{fig:deadcode}
\end{figure}

The number of projects exhibiting the dead code smell, considering all four types of dead code, is 65,760 (28.16\% of the total projects).

@@

\subsubsection{Popularity of Different Block Types}
Our analysis shows that some categories of blocks are rarely used, like the `Pen' blocks, of which only 194,885 occur within 19,090 (8.17\% of the total) programs. Hence, in future changes to the language, `Pen' blocks might be less important to users to support or maintain.

\subsubsection{Dead Code}
In our analysis, we find that more than one quarter of the Scratch projects contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, scripts that are comprised of only a \texttt{when <trigger>} block without any functionality, procedures that are not invoked, or unmatched broadcast-receive messages.

@@

\subsubsection{Exact Clones Between Sprites}
With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common. In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them. So in many cases there is no other way to share the functionality than by making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.

\subsubsection{Sharing of Scripts and Procedures}
Investigating the use of clones between projects, we observe that there are 1700 scripts that are used in multiple projects, sometimes as often as in 1600 different projects. This seems to indicate that there are common patterns in Scratch projects, which means it might be very beneficial to Scratch programmers if they could not only share their projects, but also share some of their functionality, for others to use, like a library. An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'. This might empower new Scratch users to get started quicker.

>>>>

Our study is based on data from the Scratch project repository. By scraping the list of recent projects \footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed analysis on them. To the best of our knowledge, this is the first large-scale exploratory study on Scratch projects.

@@

We obtained the set of Scratch projects by using a scraping program. Our scraping program called Kragle starts by reading the Scratch projects page\footnotemark[\ref{scratchpublic}] and thus obtaining project ids of projects that were most recently shared. Subsequently, Kragle obtains the JSON code for each of the listed projects.

@@

\subsection{Data analysis}
\label{dataAnalysis}
All scraped project data, including the list of used blocks and parameters, was imported to a relational database. We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions. This data was then imported to spreadsheets for the calculation of statistics and the generation of graphical representations. We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects, lists of which we also provide as part of the dataset.\footnotemark[\ref{repo}]

@@

Apart from whole scripts we also examined cases where scripts differed only in the first (Hat) block. This way we examine if Scratch programmers assign the same functionality to handle different types of events. Cloned functionality blocks are found to be rare: without considering the first block, only 2,243 additional clones were found in 920 projects.

The second smell that we examine is the dead code smell. We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts. Investigating the first type, we find that a significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects. This is also shown in Figure \ref{fig:procedurecalls} and it occurs in 2,079 projects. For the second type, we examined the broadcast-receive messages and found that they are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received. This lack of syncronizations occurs in 18,669 (7.99\% of the total) projects.

The third and fourth cases are detected in scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality. A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects. The majority of these scripts (86.6\%) are scripts missing the starting block. Examining the size of these dead scripts, 72.34\% are composed of a single block. As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}] The number of projects exhibiting the dead code smell, considering all four types of dead code, is 65,760 (28.16\% of the total projects).

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/12deadcode}
	\caption{Size of scripts identified as dead code}
	\label{fig:deadcode}
\end{figure}

@@

\subsubsection{Popularity of different block types}
Our analysis shows that some categories of blocks are rarely used, like the `Pen' blocks, of which only 194,885 occur within 19,090 (8.17\% of the total) programs. Hence, in future changes to the language, `Pen' blocks might be less important to users to support or maintain.

\subsubsection{Dead code}
In our analysis, we find that more than one quarter of the Scratch projects contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, scripts that are comprised of only a \texttt{when <trigger>} block without any functionality, procedures that are not invoked, or unmatched broadcast-receive messages.

@@

\subsubsection{Exact clones between sprites}
With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common. In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them. So in many cases there is no other way to share the functionality than by making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.

\subsubsection{Sharing of scripts and procedures}
Investigating the use of clones between projects, we observe that there are 1700 scripts that are used in multiple projects, sometimes as often as in 1600 different projects. This seems to indicate that there are common patterns in Scratch projects, which means it might be very beneficial to Scratch programmers if they could not only share their projects, but also share some of their functionality, for others to use, like a library. An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'. This might empower new Scratch users to get started quicker.
